---
title: "IDS investigation worksheet"
author: "by Team-Name: User1, User2, User3, User4 & User5"
date: "`r Sys.Date()`"
output: html_document
---

**Note:** You can use this file as you 'working document' where you can try out various investigation ideas and keep notes about your findings. How you use and structure this file is up to you. It is recommended that you keep notes about what you are investigating and what you find as this will make the process of creating your presentation and report easier. Please note that you _do not_ need to submit this file as part of your group project.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-lib, message = FALSE}
library(tidyverse)
library(readr)
library(dplyr)
library(janitor)
```


```{r load-data}
prison_pop_by_crime_2002_2015 <- read_csv("data/prison_pop_by_crime_2002_2015.csv", 
                                                  col_types = cols(...32 = col_skip(), 
                                                                   ...33 = col_skip(), ...34 = col_skip(), 
                                                                   `30-Jun-02` = col_number(), `30-Jun-03` = col_number(), 
                                                                   `30-Jun-04` = col_number(), `30-Jun-05` = col_number(), 
                                                                   `30-Jun-06` = col_number(), `30-Jun-07` = col_number(), 
                                                                   `30-Jun-08` = col_number(), `30-Jun-09` = col_number(), 
                                                                   `30-Jun-09(1)` = col_number(), `30-Jun-10` = col_number(), 
                                                                   `30-Sep-10` = col_number(), `31-Dec-10` = col_number(), 
                                                                   `31-Mar-11` = col_number(), `30-Jun-11` = col_number(), 
                                                                   `30-Sep-11` = col_number(), `31-Dec-11` = col_number(), 
                                                                   `31-Mar-12` = col_number(), `30-Jun-12` = col_number(), 
                                                                   `30-Sep-12` = col_number(), `31-Dec-12` = col_number(), 
                                                                   `31-Mar-13` = col_number(), `30-Jun-13` = col_number(), 
                                                                   `30-Sep-13` = col_number(), `31-Dec-13` = col_number(), 
                                                                   `31-Mar-14` = col_number(), `30-Jun-14` = col_number(), 
                                                                   `30-Sep-14` = col_number(), `31-Dec-14` = col_number(), 
                                                                   `31-Mar-15` = col_number(), `30-Jun-15` = col_number()), 
                                                  skip = 1)



prison_pop_by_crime_2015_2023 <- read_csv("data/prison_pop_by_crime_2015_2023.csv", 
                                          col_types = cols(`30-Jun-15` = col_number(), 
                                                           `30-Sep-15` = col_number(), `31-Dec-15` = col_number(), 
                                                           `31-Mar-16` = col_number(), `30-Jun-16` = col_number(), 
                                                           `30-Sep-16` = col_number(), `31-Dec-16` = col_number(), 
                                                           `31-Mar-17` = col_number(), `30-Jun-17` = col_number(), 
                                                           `30-Sep-17` = col_number(), `31-Dec-17` = col_number(), 
                                                           `31-Mar-18` = col_number(), `30-Jun-18` = col_number(), 
                                                           `30-Sep-18` = col_number(), `31-Dec-18` = col_number(), 
                                                           `31-Mar-19` = col_number(), `30-Jun-19` = col_number(), 
                                                           `30-Sep-19` = col_number(), `31-Dec-19` = col_number(), 
                                                           `31-Mar-20` = col_number(), `30-Jun-20` = col_number(), 
                                                           `30-Sep-20` = col_number(), `31-Dec-20` = col_number(), 
                                                           `31-Mar-21` = col_number(), `30-Jun-21` = col_number(), 
                                                           `30-Sep-21` = col_number(), `31-Dec-21` = col_number(), 
                                                           `31-Mar-22` = col_number(), `30-Jun-22` = col_number(), 
                                                           `30-Sep-22` = col_number(), `31-Dec-22` = col_number(), 
                                                           `31-Mar-23` = col_number(), `30-Jun-23` = col_number()), 
                                          skip = 1)




#SELECTING DATA WANTED
earliercrimes <- prison_pop_by_crime_2002_2015 %>% 
  slice(2:11)

latercrimes <- prison_pop_by_crime_2015_2023 %>% 
  slice(2:14)

# removing apare rows
earliercrimes <- earliercrimes %>% 
  select("...1" | (starts_with("30-Jun"))) %>% 
  select(-"30-Jun-09(1)")

latercrimes <- latercrimes %>% 
  select("...1" | (starts_with("30-Jun")))

#pivotting earlier data
earlierpivoted <- (earliercrimes %>% 
    pivot_longer(
      cols = c(-"...1"),
      names_to = 'date',
      values_to = "numconvicts"
    )) %>%
  pivot_wider(
    names_from = "...1",
    values_from = "numconvicts"
  )

earlierpivoted <- clean_names(earlierpivoted)

#pivoting later data
laterpivoted <- (latercrimes %>% 
                   pivot_longer(
                     cols = c(-"...1"),
                     names_to = 'date',
                     values_to = "numconvicts"
                   )) %>%
  pivot_wider(
    names_from = "...1",
    values_from = "numconvicts")

laterpivoted <- clean_names(laterpivoted)



#grouping crimes together
earlierpivotedgrouped <- earlierpivoted %>%
  mutate(
    theft_offences = (burglary + theft_and_handling),
    fraud_offences = fraud_and_forgery
  ) %>% 
  select(-"burglary", -"theft_and_handling", -"fraud_and_forgery")


laterpivotedgrouped <- laterpivoted %>% 
  mutate(
    motoring_offences = summary_non_motoring,
    other_offences = criminal_damage_and_arson + possession_of_weapons 
    + miscellaneous_crimes_against_society + summary_non_motoring + public_order_offences
  ) %>% 
  select(-"summary_non_motoring", -"criminal_damage_and_arson", -"possession_of_weapons",
         -"miscellaneous_crimes_against_society", -"summary_non_motoring", -"public_order_offences")


laterepivot <- laterpivotedgrouped %>% 
  pivot_longer(
    cols = c(-"date"),
    names_to = "crime",
    values_to = "numconvicts"
  ) %>% 
  pivot_wider(
    names_from = "date",
    values_from = "numconvicts"
  )


earlyrepivot <- earlierpivotedgrouped %>% 
  pivot_longer(
    cols = c(-"date"),
    names_to = "crime",
    values_to = "numconvicts"
  ) %>% 
  pivot_wider(
    names_from = "date",
    values_from = "numconvicts"
  )


together <-  left_join(earlyrepivot, laterepivot, by = "crime") %>% 
  select(-"30-Jun-15.y") %>% 
  rename("30-Jun-15" = "30-Jun-15.x")
view(together)

#Scatter Graph 1
#Classify crimes into two groups: violent and non-violent
violent_nonviolent <- together %>%
  mutate(
    crime_type = case_when(
      crime %in% c("violence_against_the_person", "sexual_offences", "robbery") ~ "Violent Crimes",
      crime %in% c("drug_offences", "motoring_offences", "theft_offences", "fraud_offences") ~ "Non-violent Crimes",
      crime %in% c("other_offences", "offence_not_recorded") ~ "Other",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(crime_type))

#Convert from wide to long format
violent_long <- violent_nonviolent %>%
  pivot_longer(
    cols = starts_with("30-Jun"),
    names_to = "date",
    values_to = "count"
  )

#Convert the date column to proper date format
violent_long <- violent_long %>%
  mutate(
    date = as.Date(date, format = "%d-%b-%y")
  )

#Group and sum all counts
violent_summary <- violent_long %>%
  group_by(crime_type, date) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

#Plot a scatter graph to show how different types of crime evolve in time
ggplot(violent_summary, aes(x = date, y = total_count, color = crime_type)) +
  geom_point(alpha = 0.5) +
  geom_line(size = 1.2) +
  scale_color_manual(
    values = c(
      "Violent Crimes" = "darkred",
      "Non-violent Crimes" = "orange",
      "Other" = "grey"
    )
  ) +
  labs(
    title = "Types of Crime Over Time",
    subtitle = "Grouped by Violent, Non-violent, and Other Crimes",
    x = "Date",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )
```


